The watershed transform in ITK - discussion and new developments

### 2006-04-19 article/Article.tex
Timings and connected component fix
Added timings from perf3d to the article. The connected component filter was
using long int internally, which I changed to unsigned long because the
image type used in perf3d was unsigned long.

darcs-hash:20060419045601-fafb9-9021aab8e892d96fe15c002a36efbe2d3dfaab2d.gz

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{subfigure}

@@

\tableofcontents
\section{Introduction}
The watershed transform is a tool for image segmentation that is fast
and flexible and potentially fairly parameter free. It was originally
derived from a geophysical model of rain falling on a terrain and a variety
of more formal definitions have been devised to allow development of
practical algorithms. A brief discussion of some of these definitions
and algorithms is provided in Section \ref{sect:history}.

@@

\subsubsection{Algorithm 1}
\begin{itemize}
\item lower complete the image
\item initialize the queue with all pixels except interior of minima.
\item while (!empty(queue))
   \begin{itemize}
	\item Take the pixel $p$ from the head of the queue
        \item for all pixels $q$ that are steepest upper neighbors of $p$; do
        \begin{itemize}
         \item  if $q$ is unlabeled, then

@@

There are at least two reasons why we shouldn't use this
approach. Firstly, the lower completion step is explicit, which we'd
like to avoid and secondly ALL non minima pixels are placed on the
queue, which means that the queue length is almost the same as the
number of image pixels, increasing the overhead of queue
management. We shall see that it is only necessary to have the pixels
on the borders of the regions in the queue.

@@

\subsubsection{Algorithm 2}
Using ordered queues.
\begin{itemize}
\item initialize the queue with edges of the regional minima. 
The edges are pixels that are part of a regional minima with
neighbours that are background.
\item while (!empty(queue))
   \begin{itemize}
	\item Take the pixel $p$ from the head of the queue
        \item for all pixels $q$ that are steepest upper neighbors of $p$; do
         \item  if $q$ is unlabeled then

		place $q$ in the queue

\end{itemize}

@@

\subsubsection{Algorithm 3}
Also using ordered queues.
\begin{itemize}
\item initialize the queue with background pixels that have 
      neighbors that are members of a regional minima. In practice
      this is done by checking all regional minima pixels to see
      whether they have background neighbors. The background pixels
      are placed on the queue and the label of the regional minima
      associated with them.
\item while (!empty(queue)
   \begin{itemize}
	\item Take the pixel $p$ from the head of the queue

         if $p$ is unlabelled;

           apply the associated label

         else

           check whether $p$ is a watershed pixel. $p$ is a watershed
           pixel if there is more than one label in the neighborhood.

        \item for all pixels $q$ that are steepest upper neighbors of $p$; do

	if $q$ is not in the queue;

           place in queue, associated with label($p$)

           mark as queued
   \end{itemize}
\end{itemize}

@@

Algorithm 3 requires slightly more comlexity to store label
information in the queue.

@@

@@

@@

Preventing boundary checks - not really the itk way, but if a user is
willing to throw away the image border by treating it as a watershed
line then there is no need to test for boundary conditions. There
isn't even any need to use the face calculator approach because border
pixels will be initially tagged as processed, and therefore we can be
certain their neighbors are never checked.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hyperref should be the last package to be loaded.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[dvips,
bookmarks,
bookmarksopen,
backref,
colorlinks,linkcolor={blue},citecolor={blue},urlcolor={blue},
]{hyperref}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{subfigure}
\usepackage{pseudocode}
% \usepackage{algorithm}
% \usepackage{algorithmic}

@@

\tableofcontents
\section{Introduction}

The watershed transform is a tool for image segmentation that is fast
and flexible and potentially fairly parameter free. It was originally
derived from a geophysical model of rain falling on a terrain and a variety
of more formal definitions have been devised to allow development of
practical algorithms. A brief discussion of some of these definitions
and algorithms is provided in Section \ref{sect:history}.

@@

\subsubsection{Algorithm 1}

{\em Is it really needed to present this algorithm given that there is
no lower completion ? Isn't it enough to notice somewhere that an
algorithm exist without lower completion, and which require to put
all the pixels in the queue ?}

\begin{itemize}
\item lower complete the image
\item initialize the queue with all pixels except interior of minima.
\item while (!empty(queue))
   \begin{itemize}
	\item Take the pixel $p$ from the head of the queue
        \item for all pixels $q$ that are steepest upper neighbors of $p$; do
        \begin{itemize}
         \item  if $q$ is unlabeled, then

@@

There are at least two reasons why we shouldn't use this
approach. Firstly, the lower completion step is explicit, which we'd
like to avoid and secondly {\em all} non minima pixels are placed on the
queue, which means that the queue length is almost the same as the
number of image pixels, increasing the overhead of queue
management. We shall see that it is only necessary to have the pixels
on the borders of the regions in the queue.

@@

\subsubsection{Algorithm 2}
Using ordered queues.

This algorithm does not mark the watershed lines.

@@

This algorithm take a grayscale image $I$ and a marker image $M$
which contains a set of markers with different labels. It can be
splitted in to steps:
\begin{enumerate}
  \item copy the marker image $M$ to the output image, and put
all the marker pixels $p$ with background neighbors in the
ordered queue, with the priority $I[p]$.
  \item get a pixel from the ordered queue at the highest
priority level, and propagate the label of $p$ to all the unlabeled
neighbors of $p$. The unlabeled neighbors $q$ are added in the
ordered queue with priority $I[q]$, if $I[q]$ is greater than
the priority level of $p$, or with the priority of $p$ otherwise.
\end{enumerate}

The detailed algorithm is shown in Figure \ref{BeucherAlgorithm}.

\begin{figure}[htbp]
\centering
\small
\begin{pseudocode}[framebox]{BeucherWatershed}{I,M}
\COMMENT{initialization stage} \\
queue \GETS \emptyset \\
\FOREACH p \in M \DO
\BEGIN
  \COMMENT{copy the marker image to the output image}\\
  O[p] \GETS M[p] \\
  \COMMENT{put the edges of the marker in the queue}\\
  \IF M[p] \neq WSLABEL
  \THEN
  \BEGIN
    bgNeighbor \GETS \FALSE \\
    \FOREACH q \in \CALL{N}{p} \DO 
    \BEGIN
      \IF M[q] = WSLABEL
        \THEN bgNeighbor \GETS \TRUE \\
    \END\\
    \IF bgNeighbor \STMTNUM{5em}{algo:Beucher:bgNeighbor}
      \THEN
      queue[I[p]] \GETS queue[I[p]] \cup \{p\} \\
  \END
\END
\\
\\
\COMMENT{flooding stage} \\
\WHILE queue \neq \emptyset \DO
\BEGIN
  v \GETS \CALL{first}{queue}\\
  \WHILE queue[v] \neq \emptyset \DO
  \BEGIN
    p \GETS \CALL{first}{queue[v]} \\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
    \IF O[q] = WSLABEL
    \THEN
    \BEGIN
      O[q] \GETS O[p] \\
      v \GETS \CALL{MAX}{I[q], v} \STMTNUM{5em}{algo:Beucher:max} \\
      queue[v] \GETS queue[v] \cup \{q\} \\
    \END
    \END
  \END
\END
\end{pseudocode}
\caption{Simple watershed algorithm.\label{BeucherAlgorithm} $FIRST(queue)$ extract the highest priority element from the queue. $N(p)$ return the list of p's neighbors. $MAX(a,b)$ returns the maximum of $a$, and $b$. $WSLABEL$ is the label used to define background and watershed pixels (usally $0$). The lower completion is done by the statement (\ref{algo:Beucher:max}). Determine if a marker pixel before adding it in the queue (\ref{algo:Beucher:bgNeighbor}) is not required, but helps to decrease memory usage in case of large markers.}
\end{figure}

% \begin{itemize}
% \item initialize the queue with edges of the regional minima. 
% The edges are pixels that are part of a regional minima with
% neighbours that are background.
% 
% \item while (!empty(queue))
%    \begin{itemize}
% 	\item Take the pixel $p$ from the head of the queue
%         \item for all pixels $q$ that are steepest upper neighbors of $p$; do
%          \item  if $q$ is unlabeled then 
% 
%      		propagate label$(p)$ to label$(q)$;
% 	
% 		place $q$ in the queue
% 
%                 else if (label$(q) \ne \mbox{WS})$ and  (label$(q) \ne \mbox{label}(p)$ then
% 
%                 label($q$) = WS;
% 	\end{itemize}
% 
% \end{itemize}

\subsubsection{Algorithm 3}
Also using ordered queues.

This algorithm take a grayscale image $I$ and a marker image $M$
which contains a set of markers with different labels. It can be
splitted in to steps:
\begin{enumerate}
  \item copy the marker image $M$ to the output image, and put
all the background pixels $p$ with markers in there neighbors in the
ordered queue, with the priority $I[p]$, and mark $p$ as already in
the queue. All the marker pixels are marked as already in the queue.
  \item get a pixel $p$ from the ordered queue at the highest
priority level. If the neighborhood contains only the same label
or watershed pixels, mark $p$ with that label. The neighbors $q$ 
of $p$ which are not already in the queue are added to the queue 
with priority $I[q]$, if $I[q]$ is greater than the priority level
of $p$, or with the priority of $p$ otherwise.
\end{enumerate}

The detailed algorithm is shown in Figure \ref{MeyerAlgorithm}.

Contrary to the previous algorithm, the label in the output image
can't be used to determine if the pixel has already been added to
the queue, and so we need to maintain this status by another way.
It is the role of $S$ in Figure \ref{MeyerAlgorithm}. We'll discuss
in \ref{sect:implementation} how this status can be implemented.

This algorithm mark the watershed lines.

It must be noticed that this algorithm never propagate watershed
lines, and so produce very thin watershed lines.

\begin{figure}[htbp]
\centering
\small
\begin{pseudocode}[framebox]{MeyerWatershed}{I,M}
\COMMENT{initialization stage} \\
\FOREACH p \in M \DO
  S[p] \GETS \FALSE \\
queue \GETS \emptyset \\
\FOREACH p \in M \DO
\BEGIN
  \COMMENT{copy the marker image to the output image}\\
  O[p] \GETS M[p] \\
  \IF M[p] \neq WSLABEL
  \THEN
  \BEGIN
    S[p] \GETS \TRUE \\
    \COMMENT{add all the neighbors of $p$ with WSLABEL to the queue}\\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
      \IF \neg S[q] \AND M[q] = WSLABEL 
      \THEN
      \BEGIN
        S[q] \GETS \TRUE \\
        queue[I[q]] \GETS queue[I[q]] \cup \{q\} \\
      \END
    \END
  \END
\END
\\
\\
\COMMENT{flooding stage} \\
\WHILE queue \neq \emptyset \DO
\BEGIN
  v \GETS \CALL{first}{queue}\\
  \WHILE queue[v] \neq \emptyset \DO
  \BEGIN
    p \GETS \CALL{first}{queue[v]} \\
    \COMMENT{found the label of $p$ and determine if it is a watershed pixel} \\
    label \GETS WSLABEL \\
    watershed \GETS \FALSE \\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
      \IF O[q] \neq WSLABEL \AND \neg watershed
      \THEN
        \BEGIN
        \IF label \neq WSLABEL \AND O[q] \neq label \\
        \THEN
          watershed \GETS \TRUE \\
        \ELSE
          label \GETS O[q] \\
        \END
    \END \\
    \COMMENT{if $p$ is a watershed pixel, don't propagate this pixel} \\
    \IF \neg watershed
    \THEN
      \BEGIN
        \COMMENT{assign the label value to the current pixel} \\
        O[p] \GETS label \\
        \COMMENT{put all the neighbors of p in the queue} \\
        \FOREACH q \in \CALL{N}{p} \DO
        \BEGIN
          \IF \neg S[q]
          \THEN
            \BEGIN
            S[q] \GETS \TRUE \\
            h \GETS \CALL{MAX}{I[q], v} \STMTNUM{5em}{algo:Meyer:max} \\
            queue[h] \GETS queue[h] \cup \{q\} \\
            \END
        \END
      \END
  \END
\END
\end{pseudocode}
\caption{Watershed algorithm with watershed lines.\label{MeyerAlgorithm} $FIRST(queue)$ extract the highest priority element from the queue. $N(p)$ return the list of p's neighbors. $MAX(a,b)$ returns the maximum of $a$, and $b$. $WSLABEL$ is the label used to define background and watershed pixels (usally $0$). The lower completion is done by the statement (\ref{algo:Meyer:max})}
\end{figure}

% \begin{itemize}
% \item initialize the queue with background pixels that have 
%       neighbors that are members of a regional minima. In practice
%       this is done by checking all regional minima pixels to see
%       whether they have background neighbors. The background pixels
%       are placed on the queue and the label of the regional minima
%       associated with them.
% \item while (!empty(queue)
%    \begin{itemize}
% 	\item Take the pixel $p$ from the head of the queue
% 	
%          if $p$ is unlabelled;
% 
%            apply the associated label
% 
%          else
%  
%            check whether $p$ is a watershed pixel. $p$ is a watershed
%            pixel if there is more than one label in the neighborhood.
% 
%         \item for all pixels $q$ that are steepest upper neighbors of $p$; do
% 	
% 	if $q$ is not in the queue;
% 
%            place in queue, associated with label($p$)
%         
%            mark as queued
%    \end{itemize}
% \end{itemize}

@@

Algorithm 3 requires slightly more complexity to store label
information in the queue.

@@

The new work implements algorithms 2 and 3 -- algorithm 2 is used when
the watershed lines are not marked while algorithm 3 is used when they
are.

@@

\section{Implementation}
\label{sect:implementation}

We provide with this article two new classes:

  \subsection{MorphologicalWatershedFromMarkersImageFilter}

\verb$MorphologicalWatershedFromMarkersImageFilter$ implement the
algorithms \ref{sect:BeucherAlgorithm} and \ref{sect:MeyerAlgorithm}.
They are accessible respectively when the attribute \verb$MarkWatershed$
is \verb$false$ or is \verb$true$. The filter is templated over the
input image type, which can be any scalar type, and the label image type
which should be an integer. The filter take 2 inputs:
\begin{itemize}
  \item a grayscale image which can be set with \verb$SetInput(image)$,
\verb$SetInput(0, image)$ or \verb$SetInput1(image)$.
  \item a marker image which can be set with \verb$SetMarkerImage(image)$
and \verb$SetInput2(image)$. \verb$SetInput(0, image)$ should be avoided
because it will fail if the input image and the marker image are not of the
same type. The markers are all the non-zero values in the image.
\end{itemize}

The algorithm \ref{sect:MeyerAlgorithm} require to keep the status of all
the pixels: {\em already added to the queue} or {\em not yet added to the queue}.
This task is often done by using negative values in the output image
for the pixels not yet added to the queue. In our implementation, we have
chosen to use a binary image to store this, while labels always take values
above 0 and watershed lines are marked with value 0. The negative values
option have several drawbacks:
\begin{itemize}
  \item it requires to use a different image type than the label type defined
by the user in case the user have chosen an unsigned pixel type.
  \item it requires to use one of the value which can be possibly aready used
by a label in the marker image, in case of a signed type.
\end{itemize}
Those problems can be avoided\footnote{It's not true for the unsigned long
pixel type} by using a type which contains a larger range of value than the
type choosed by the user, but then, we must copy the data to an image of
the type chosen by the user at the end of the algorithm.

We prefered using a separated binary image (of type \verb$itk::Image<bool, dim>$)
to store this information. This solution avoid the need to copy the image at
the end, and so should be more efficient in computation time, and require only
one bit per pixel and so should be more memory efficient as well.

The connectivity is defined using the usual ITK way: \verb$SetFullConnected(false)$
(4-connected in 2D, 6-connected in 3D) or \verb$SetFullConnected(true)$ (8-connected
in 2D, 26-connected in 3D)\footnote{One should regret to not have the 18-connected
connectivity in 3D}.

  \subsection{MorphologicalWatershedImageFilter}

\verb$MorphologicalWatershedImageFilter$ perform a watershed transformation
without marker. It encapsulate several filters:
\begin{itemize}
  \item \verb$HMinimaImageFilter$
  \item \verb$RegionalMinimaImageFilter$
  \item \verb$ConnectedComponentImageFilter$
  \item \verb$MorphologicalWatershedFromMarkersImageFilter$
\end{itemize}
While \verb$RegionalMinimaImageFilter$, \verb$ConnectedComponentImageFilter$,
and \verb$MorphologicalWatershedFromMarkersImageFilter$ are the usal steps
of the waterhed usage, the \verb$HMinimaImageFilter$ is a convenient way
to avoid the classical watershed's over segmentation by allowing the user
to select the dynamic of the minima. The minimum dynamic of minima can be
selected with \verb$SetThreshold()$.

\subsection{Other performance issues/comments}
\begin{itemize}
\item Preventing boundary checks - not really the itk way, but if a user is
willing to throw away the image border by treating it as a watershed
line then there is no need to test for boundary conditions. There
isn't even any need to use the face calculator approach because border
pixels will be initially tagged as processed, and therefore we can be
certain their neighbors are never checked.
\item Streaming and threading  - queue based algorithms are generally difficult 
to thread efficiently and generally need to have all of the input
available so prevent streaming from working well. The existing ITK
watershed does provide a multithreaded implementation and similar
techniques may be possible with the new implementation. Other
possibilities include using threads to improve performance of queues.
\end{itemize}

@@

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=4]{threshold}}
\subfigure[No minima threshold]{\includegraphics[scale=4]{threshold00-rgb}}
\subfigure[Minima threshold = 10]{\includegraphics[scale=4]{threshold10-rgb}}
\subfigure[Minima threshold = 20]{\includegraphics[scale=4]{threshold20-rgb}}
\subfigure[Minima threshold = 30 and Minima threshold = 40]{\includegraphics[scale=4]{threshold30-rgb}}
\subfigure[Minima threshold = 50]{\includegraphics[scale=4]{threshold50-rgb}}
\subfigure[markers]{\includegraphics[scale=4]{threshold-markers}}
\subfigure[Watershed from markers]{\includegraphics[scale=4]{threshold-markersMf-rgb}}
\caption{Different results of watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image. All the watersheds were performed with $MarkWatershed = true$ and $FullyConnected = false$.}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=4]{threshold}}
\subfigure[No minima threshold]{\includegraphics[scale=4]{threshold00-itk-rgb}}
\subfigure[Minima threshold = 10]{\includegraphics[scale=4]{threshold10-itk-rgb}}
% \subfigure[Minima threshold = 20]{\includegraphics[scale=4]{threshold20-itk-rgb}}
\subfigure[Minima threshold = 40]{\includegraphics[scale=4]{threshold40-itk-rgb}}
\subfigure[Minima threshold = 63]{\includegraphics[scale=4]{threshold63-itk-rgb}}
\subfigure[Minima threshold = 65]{\includegraphics[scale=4]{threshold65-itk-rgb}}
\subfigure[markers]{\includegraphics[scale=4]{threshold-markers}}
\subfigure[Watershed from markers]{\includegraphics[scale=4]{threshold-markers-itk-rgb}}
\caption{Different results of watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image..}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\subfigure[]{\includegraphics[scale=10]{pass-values}}
\subfigure[]{\includegraphics[scale=10]{pass-values-mf-rgb}}
\subfigure[]{\includegraphics[scale=10]{pass-values-mF-rgb}}
\subfigure[]{\includegraphics[scale=10]{pass-values-Mf-rgb}}
\subfigure[]{\includegraphics[scale=10]{pass-values-MF-rgb}}
\subfigure[]{\includegraphics[scale=10]{pass-values-itk-rgb}}

\subfigure[]{\includegraphics[scale=4]{plateau}}
\subfigure[]{\includegraphics[scale=4]{plateau-mf-rgb}}
\subfigure[]{\includegraphics[scale=4]{plateau-mF-rgb}}
\subfigure[]{\includegraphics[scale=4]{plateau-Mf-rgb}}
\subfigure[]{\includegraphics[scale=4]{plateau-MF-rgb}}
\subfigure[]{\includegraphics[scale=4]{plateau-itk-rgb}}

\subfigure[]{\includegraphics[scale=10]{thick-lines}}
\subfigure[]{\includegraphics[scale=10]{thick-lines-mf-rgb}}
\subfigure[]{\includegraphics[scale=10]{thick-lines-mF-rgb}}
\subfigure[]{\includegraphics[scale=10]{thick-lines-Mf-rgb}}
\subfigure[]{\includegraphics[scale=10]{thick-lines-MF-rgb}}
\subfigure[]{\includegraphics[scale=10]{thick-lines-itk-rgb}}

\subfigure[]{\includegraphics[scale=6]{button-hole}}
\subfigure[]{\includegraphics[scale=6]{button-hole-mf-rgb}}
\subfigure[]{\includegraphics[scale=6]{button-hole-mF-rgb}}
\subfigure[]{\includegraphics[scale=6]{button-hole-Mf-rgb}}
\subfigure[]{\includegraphics[scale=6]{button-hole-MF-rgb}}
\subfigure[]{\includegraphics[scale=6]{button-hole-itk-rgb}}

\caption{}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\subfigure[]{\includegraphics[scale=0.5]{cthead1}}
\subfigure[]{\includegraphics[scale=0.5]{cthead1-markers}}
\subfigure[]{\includegraphics[scale=0.5]{cthead1mf-rgb}}
\subfigure[]{\includegraphics[scale=0.5]{cthead1mF-rgb}}
\subfigure[]{\includegraphics[scale=0.5]{cthead1Mf-rgb}}
\subfigure[]{\includegraphics[scale=0.5]{cthead1MF-rgb}}
\subfigure[]{\includegraphics[scale=0.5]{cthead1itk-rgb}}
\caption{}
\end{center}
\end{figure}
